Evaluating \Htwo{} and \COtwo{} Network Strategies for the European Energy System

### 2023-12-20 report/submissions/journal/paper/paper.tex
paper: tidy up
incorp reviews, restructure figures

% alternative
\newcommand{\carbon}{CO$_2$}
\newcommand{\hydrogen}{H$_2$}
\newcommand{\carbongrid}{CO$_2$\=/Grid}
\newcommand{\hydrogengrid}{H$_2$\=/Grid}
\newcommand{\baselinemodel}{Baseline model}
\newcommand{\carbonmodel}{CO$_2$\=/Grid model}
\newcommand{\hydrogenmodel}{H$_2$\=/Grid model}
\newcommand{\hybridmodel}{Hybrid model}

@@

\author{
    Fabian Hofmann, Christoph Tries, Fabian Neumann, Lisa Zeyen, Tom Brown \\
    \textit{Institute of Energy Technology} \\
    \textit{Technical University of Berlin}\\
    Berlin, Germany \\
    m.hofmann@tu-berlin.de
}

@@

\begin{abstract}
    Hydrogen and carbon dioxide transportation are considered crucial in climate-neutral energy systems, with hydrogen transport enabling energy imports to high-demand areas and carbon transport aiding in emissions export from high-emission areas. Yet, possible synergies and competitions between the two systems are not fully understood. Our study employs optimization techniques to develop a cost-optimal European energy system, integrating production, transport and storage of both carbon dioxide and hydrogen, along with renewable energy sources. Results indicate that, in direct comparison, a hydrogen network is more cost-effective than a carbon network, facilitating low-cost hydrogen transport to Central Europe and local carbon capture and utilization. However, in a hybrid scenario, the carbon network effectively complements the hydrogen network, promoting carbon capture from biomass and reducing reliance on direct air capture. This research demonstrates the effectiveness of combining hydrogen and carbon transport networks with power grid expansion in achieving climate neutrality in Europe.
\end{abstract}

@@

The transition to a carbon-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both \COtwo{} and \Htwo{}, also referred to as carbon and hydrogen, is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including of carbon capture, transport, storage, utilization, and sequestration. Hydrogen offers an efficient way to transport and store energy over long distances and timeframes, and can be produced from electrolysis or steam methane reforming (SMR), for example. Carbon can be stored in geological formations to permanently extract it from the atmosphere, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels as for example needed for aviation in form of synthetic kerosene from Fischer-Tropsch synthesis, for shipping in form of synthetic methanol and for industrial feedstocks through synthetic methane. From hereon, we jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Both \COtwo{} and \Htwo{} networks are likely to play a crucial role in the cost-effective integration of these technologies to enable net-zero economies.

Recently, policymakers and industry in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing on the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established with dedicated instruments supporting both the ramp-up of the hydrogen and carbon economy~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier. Some natural gas pipelines are already repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synthetic fuels on the one hand and offer \COtwo{} offtake on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \COtwo{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it's important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily for enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

However, up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or replace each other. Both can bridge the misalignment of sources and sinks for carbon and hydrogen which may well be determined by exogenous factors. A hydrogen network can supply regions with geographically fixed hydrogen demand, such as for steelmaking, with hydrogen from regions with the best renewable resources as well as enable CU at the site of point-source CC. On the other hand, the carbon transport system allows for transporting captured carbon to regions with sequestration potentials or high-quality renewable resources, the latter enabling cost-effective CU.
In the literature, the two network approaches and underlying technologies have been discussed in a number of publications, all of which, however, dealt with the isolated effects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks which are crucial to provide a holistic view of the energy system and its technological interactions. A remarkable example was developed by Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022}, which comprehensively demonstrates the interchangeability of hydrogen and electricity grid expansion in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses most relevant sectors, however neglects the option to transport carbon across regions.
The publication by Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \COtwo{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects the sectors that will need to handle most \COtwo{} in the future.
A comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \COtwo{} transport system in Switzerland, connecting to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and co-benefits from connecting the \COtwo{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \COtwo{} transport. Euro-Calliope, for example, does neither feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region~\cite{pickeringDiversityOptionsEliminate2022}.

To our knowledge, no study has yet considered the co-optimization and comprehensive assessment of both \COtwo{} and \Htwo{} networks in a fully sector-coupled energy system. However, we argue that such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model which optimizes operation and investments in all relevant energy sectors. To investigate the competing transport dynamics between \COtwo{} and \Htwo{} networks, we contrast the following four models that have none of these networks, only one, or both combined:
%
\begin{itemize}
    \item[] \textit{Baseline:} Neither \COtwo{} nor \Htwo{} networks are constructed.
    \item[] \textit{\COgrid{}:} Only the \COtwo{} network is developed.
    \item[] \textit{\Hgrid{}:} Only the \Htwo{} network is developed.
    \item[] \textit{Hybrid:} Both \COtwo{} and \Htwo{} networks are developed.
\end{itemize}
%
In anticipation of the future \carbon{} emission targets, the models are subjected to the following two scenarios:
\begin{itemize}
    \item[] \textit{Net\=/Zero}: aligning with the EU's 2050 emission targets. Limiting carbon sequestration to 200~Mt/a. This scenario is also denoted as NZ.
    \item[] \textit{Net\=/Negative}: 10\% net-negative emissions relative to 1990 levels, equaling 460~Mt \COtwo{} annually. Limiting carbon sequestration to 660~Mt/a. Also denoted as NN.
\end{itemize}

We analyze how the deployment of carbon and hydrogen networks affects different decarbonization strategies in the models and scenarios and how investments in none or only one of the transport technologies lead to inefficient market layouts. We do this by first focusing on the net-zero scenario and a detailed description of the different technological and geographical impacts of the transportation systems. Subsequently our analysis demonstrates the extent to which different deployment strategies remain robust under a tightened emissions target in the Net-Negative scenario.

@@

Focusing first on the Net-Zero scenario, we can highlight notable characteristics of the energy system buildouts shared by all models, as well as significant variations in system costs and technology deployment between the models (see Fig.~\ref{fig:cost_bar}).

@@

Total annual system costs in the \modBase{} are highest, totalling \label{}779~bn€. The models with additional transport options for carbon and hydrogen achieve a more efficient system allocation and lower total system costs, with \label{}3.1\% cost reduction (\label{}754~bn€) in the \modCO{}, and \label{}4.8\% (\label{}742~bn€) in the \modH{}, and \label{}5.2\% (\label{}738~bn€) in the the \modHybrid{}.

In all models, almost half of the system costs are spent on primary electricity production from wind and solar power, with around another \label{}10\% spent on hydro, nuclear and biomass. For the \baselinemodel, transmission system costs for electricity, gas, hydrogen and carbon are below \label{}4\% (\label{}25~bn€), with the electricity grid making up more of the system costs than the other three networks combined. This is not the case for the other models, where combined pipeline costs slightly surpass electricity transmission system costs (up to 20\% for the \hybridmodel). Capital expenditures on carbonaceous fuel production are around \label{}4\% (\label{}30~bn€) in each model, while electrolysis makes up around \label{}4.7\% (\label{}37~bn€). Heat pump installations account for around \label{}8\% of total system costs (64~bn€ in the \baselinemodel).

System cost differences between the models are driven by renewable energy resources (solar, wind, and bioenergy), the \COtwo{} and \Htwo{} networks, as well as carbon capture technologies (DAC and CC at point sources).
As we show below, these are due to substantial, albeit fundamentally different, gains in flexibility and a shift in carbon utilization and sequestration strategies by the \Htwo{} and \COtwo{} networks.

\subsection*{Transport systems unlock point\=/source carbon capture}

In comparison to the other three models, the \modBase{} reveals significant inefficiencies: On the one hand, regions with no access to sequestration sites (i.e., ``landlocked'' regions) underutilize CC potentials from process emissions, fossil gas and biomass. On the other hand, sequestration regions rely on costly DAC on-site. Furthermore, hydrogen and power prices show high regional differences, indicating system inefficiencies due to transport constraints caused by non\=/existent hydrogen or insufficient power transmission. Prices are significantly higher in Central Europe, a region with high population density and high concentration of energy-intensive industries.

By enabling the efficient transportation of carbon, hydrogen, or both, these models reduce the reliance on high-cost DAC, and source more carbon from low-cost point-source CC on biomass and gas in industrial applications (see Fig.~\ref{fig:balance_captured_carbon}). The \modH{} only captures \label{}193~Mt of \COtwo{} from DAC compared to \label{}392~Mt in the \modBase{}. Instead, the \modH{} captures more carbon from process emissions as well as gas and biomass for industry. The \modCO{} uses only \label{}154~Mt from DAC, and instead also captures more carbon than the \modH{} from gas for industry and biomass CHPs. And finally, the \modHybrid{} captures only \label{}112~Mt from DAC, but slightly more carbon from all other CC technologies.
Also, the transport of \carbon{} and \hydrogen{} reduce the reliance on methanation and therefore the overall demand for CC volumes in the model. Another knock-on effect of less reliance on DAC is its reduced endogenous demand for dedicated renewables as well as heat pumps and biomass CHPs to supply power and heat.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/balance_bar_carbon.png}
    \caption{Balance of captured carbon for all models in the Net-Zero scenario. Positive values indicate carbon capture, negative values indicate carbon consumption. By integrating \Htwo{} and \COtwo{} networks, the predominant method for carbon removal shifts from Direct Air Capture (DAC) to a bioenergetic process that incorporates carbon capture. At the same time the reliance on methanation decreases.}
    \label{fig:balance_captured_carbon}
\end{figure}

The increase in captured carbon from point sources in the models with \carbon{} and/or \hydrogen{} transport is more related to an upgrading of existing industrial or power plant facilities to capture carbon rather than a shift in technologies. This is evident when analyzing the capture share of the different CC technologies (see Fig.~\ref{fig:captureshare_line}).

>>>>

% alternative
\newcommand{\carbon}{CO$_2$}
\newcommand{\hydrogen}{H$_2$}
\newcommand{\carbongrid}{CO$_2$\=/Grid}
\newcommand{\hydrogengrid}{H$_2$\=/Grid}
\newcommand{\baselinemodel}{Baseline model}
\newcommand{\carbonmodel}{CO$_2$\=/Grid model}
\newcommand{\hydrogenmodel}{H$_2$\=/Grid model}
\newcommand{\hybridmodel}{Hybrid model}
\newcommand{\baselinescenario}{Baseline scenario}
\newcommand{\carbonscenario}{CO$_2$\=/Grid scenario}
\newcommand{\hydrogenscenario}{H$_2$\=/Grid scenario}
\newcommand{\hybridscenario}{Hybrid scenario}

@@

\author{
    Fabian Hofmann, Christoph Tries, Fabian Neumann, Elisabeth Zeyen, Tom Brown \\
    \textit{Institute of Energy Technology} \\
    \textit{Technical University of Berlin}\\
    Berlin, Germany \\
    m.hofmann@tu-berlin.de
}

@@

\begin{abstract}
    Hydrogen and carbon dioxide transport can both play an important role in climate-neutral energy systems. Hydrogen networks help serve regions with high energy demand, while excess emissions are transported away in carbon dioxide networks. When it comes to the synthesis of carbonaceous fuels, it is less clear which input should be transported: hydrogen to carbon point sources or carbon to low-cost hydrogen. We explore the potential synergies and competition of both networks in a cost-optimal carbon-neutral European energy system. In direct comparison, a hydrogen network is more cost-effective than a carbon network, since it serves to transport hydrogen to demand and to point source of carbon for utilization. However, in a hybrid scenario where both networks are present, the carbon network effectively complements the hydrogen network, promoting carbon capture from biomass and reducing reliance on direct air capture. This study shows how the European energy system in future net-zero and net-negative emissions scenarios can benefits from integrating both networks.
\end{abstract}

@@

The transition to a climate-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both carbon dioxide (\COtwo{}) and hydrogen (\Htwo{}) is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including carbon capture, transport, storage, utilization, and sequestration. Hydrogen is consider in several sectors in industry not amenable to electrification, and offers an efficient way to transport and store energy over long distances and timeframes, and can be produced from electrolysis or steam methane reforming (SMR), for example. Captured carbon can be stored in geological formations to permanently extract it from the atmosphere, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels to supply for example synthetic kerosene for aviation, synthetic methanol for shipping or synthetic methane for industrial feedstocks. From hereon, we jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Both \COtwo{} and \Htwo{} networks are likely to play a crucial role in the cost-effective integration of these technologies to enable net-zero economies.

Recently, policymakers and industry in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing to the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established with dedicated instruments supporting both the ramp-up of the hydrogen and carbon economy~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier. Some natural gas pipelines have already been repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synthetic fuels on the one hand and offer \COtwo{} offtake on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \COtwo{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it's important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily to supply \carbon{} to enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

Up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or even compete with each other. Both can bridge the misalignment of sources and sinks for carbon and hydrogen. A hydrogen network can supply regions with geographically fixed hydrogen demand, such as for steelmaking, with hydrogen from regions with the best renewable resources as well as enable CU at the site of point-source CC. On the other hand, the carbon transport system allows for transporting captured carbon to regions with sequestration potentials or high-quality renewable resources, the latter enabling cost-effective CU.

In the literature, the two network approaches and underlying technologies have been discussed in several of publications, all of which, however, dealt with the isolated aspects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks which are crucial to provide a holistic view of the energy system and its technological interactions. Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022} demonstrated the interchangeability of hydrogen and electricity grid expansion in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses most relevant sectors, however it neglects the option to transport carbon across regions.
Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \COtwo{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects the sectors like industry that will need to handle most \COtwo{} in the future.
A comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \COtwo{} transport system in Switzerland, connecting to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and co-benefits from connecting the \COtwo{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \COtwo{} transport. Euro-Calliope, for example, does neither feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region~\cite{pickeringDiversityOptionsEliminate2022}.

No study has yet considered the co-optimization and comprehensive assessment of both \COtwo{} and \Htwo{} networks in a fully sector-coupled energy system. However, such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model which optimizes operation and investments in all relevant energy sectors to supply projected energy demands for 2050 (see Fig.~\ref{fig:total-demand-bar}). It is assumed that Europe is self-sufficient in energy and does not import any fuels. This drives carbonaceous fuels production to be located at sites where both \carbon{} and \hydrogen{} can be provided at low cost.

\begin{figure}[h!]
    \includegraphics[width=\linewidth]{baseline/figures/90_nodes/demand_bar.png}
    \caption{Assumptions on exogenous demand, derived from~\cite{piamanzGeoreferencedIndustrialSites2018,muehlenpfordtTimeSeries2019,mantzosJRCIDEES20152018,NationalEmissionsReported2023,EurostatCompleteEnergyBalance,uwekrienDemandlib2023}. The figure shows the total annual energy demand for each energy source which determine the model's endogenous investments and operation. Endogenous processes can lead to higher total production volumes of some energy carriers, e.g., the demand for methanol requires more hydrogen and carbon as secondary (energy) inputs, which are not considered here. The model are defined per region and time stamp.}
    % TODO: adjust labels to show that these are exogenous assumptions
    \label{fig:total-demand-bar}
\end{figure}

% TODO: This drives all CCU....
% TODO: The demands should be mentioned properly introduced

To investigate the competing transport dynamics between \COtwo{} and \Htwo{} networks, we contrast the following four model scenarios:
%
\begin{itemize}
    \item[] \textit{Baseline:} Neither \COtwo{} nor \Htwo{} networks are constructed.
    \item[] \textit{\COgrid{}:} Only the \COtwo{} network is developed.
    \item[] \textit{\Hgrid{}:} Only the \Htwo{} network is developed.
    \item[] \textit{Hybrid:} Both \COtwo{} and \Htwo{} networks are developed.
\end{itemize}
%
% In anticipation of the future \carbon{} emission targets, the models are subjected to the following two scenarios:
% \begin{itemize}
%     \item[] \textit{Net\=/Zero}: aligning with the EU's 2050 emission targets. Limiting carbon sequestration to 200~Mt/a. This scenario is also denoted as NZ.
%     \item[] \textit{Net\=/Negative}: 10\% net-negative emissions relative to 1990 levels, equaling 460~Mt \COtwo{} annually. Limiting carbon sequestration to 660~Mt/a. Also denoted as NN.
% \end{itemize}

We analyze how the deployment of carbon and hydrogen networks affects different decarbonization strategies in the models and scenarios and how investments in none or only one of the transport technologies lead to inefficient market layouts. We do this by focusing first on a net-zero \carbon{} emissions target, while limiting carbon sequestration to a total 200~Mt/a. We give a detailed description of the different technological and geographical impacts of the transportation systems. Subsequently, our analysis demonstrates the extent to which different deployment strategies remain robust under a tightened net-negative \carbon{} emissions target.

@@

All scenarios share key energy system characteristics but differ in system costs and technology use (Fig.~\ref{fig:cost_bar}). The \modBase{} has the highest annual cost at 779~bn€/a. Models with \carbon{} and \hydrogen{} transport options are more cost-effective, reducing costs by 3.1\% in \modCO{}, 4.8\% in \modH{}, and 5.2\% in \modHybrid{}.
% in system costs: 779bn€ (baseline), 754bn€ (carbon), 742bn€ (hydrogen), 738bn€ (hybrid)

@@

% In every model, nearly half the costs are attributable to wind and solar electricity production, and about 10\% to hydro, nuclear, and biomass. In the \baselinemodel, transmission costs (electricity, gas, hydrogen, carbon) are below 4\% (25 bn€) and are dominated by electricity grid expenses. In contrast, the other models have higher combined pipeline costs, peaking at 20\% in the \hybridmodel. Both carbonaceous fuel production and electrolysis account for about 4\% (30 bn€) and 4.7\% (37 bn€) of costs respectively, with heat pump installations contributing around 8\% (64 bn€ in the \baselinemodel).

System cost differences between the models are driven by wind and solar as well as carbon capture technologies. In scenarios with \carbon{} and \hydrogen{} transportation, the model primarily reduces the reliance on high-cost DAC from 392~Mt/a carbon removal in the \baselinescenario{} to 112~Mt/a in the \hybridscenario{} (see Fig.~\ref{fig:balance_captured_carbon}). At the same time, carbon capture from biomass combustion, gas-based industry processes and biogas-to-gas upgrading increase. The increased usage of upgraded biogas leads to a lower synthetic methane production.
% At the same time, the transport of \carbon{} and \hydrogen{} reduce the reliance on methanation and therefore the overall demand for CC volumes in the model. Another knock-on effect of less reliance on DAC is its reduced endogenous demand for dedicated renewables as well as heat pumps and biomass CHPs to supply power and heat.
\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/balance_bar_carbon.png}
    \caption{Balance of captured carbon for all models in the Net-Zero scenario. Positive values indicate carbon capture, negative values indicate carbon consumption. By integrating \Htwo{} and \COtwo{} networks, the predominant method for carbon removal shifts from Direct Air Capture (DAC) to a bioenergetic process that incorporates carbon capture. At the same time the reliance on methanation decreases.}
    \label{fig:balance_captured_carbon}
\end{figure}

In the \baselinescenario{}, where there is no carbon and hydrogen transport, hydrogen must be produced where it is consumed. For CU, either \hydrogen{} must be produced where \carbon{} is cheap at point sources, or DAC must be used at sites with low-cost \hydrogen{}. Where \carbon{} from point sources is not able to be sequestered, it has to be compensated elsewhere by negative emissions from DAC or biomass capture with sequestration. All of these options are sub-optimal.

\subsection*{Networks unlock low-cost \hydrogen{} and \carbon{}}

Both a carbon network and a hydrogen network can help alleviate these inefficiencies. A carbon network allows \carbon{} from point sources to be transported to low-cost hydrogen as well as sequestration sites. A hydrogen network on the other hand allows low-cost hydrogen to be transported from regions with good renewable resources to point sources for CU. This results in fundamentally different network flows in each scenario (see Fig.~\ref{fig:balance_map}).

\begin{figure*}[ht!]
    \centering
    \includegraphics[width=0.98\linewidth]{comparison/single-technologies/figures/90_nodes/balance_map_dedicated.png}
    \caption{Average production, consumption, flows and prices of the carbon (top line) and hydrogen (bottom line) sectors in the \carbongrid{} (left) and the \hydrogenscenario{} (right). For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Lines and arrows show the interregional transportation. Carbon sequestration offshore are drawn as full circles.
    }
    \label{fig:balance_map}
\end{figure*}

In the \carbonscenario, there are three major transport routes of the low-cost carbon captured in Central and Eastern Europe (see Fig.~\ref{fig:balance_map}, top left). A large part of the carbon is transported from Western Europe to the Iberian Peninsula (8~kt/h), the British Isles (7.5~kt/h) and Denmark (2.8~kt/h), where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. A second part is transported to methanation plants in Italy and Greece (2.8~kt/h), where the synthetic methane produced is fed into the gas grid in the vicinity of major gas import hubs. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. Almost none of the captured carbon is used locally at the point-source. In addition to electrolysis in regions with the good renewable resources to supply CU ($\sim$125~GW$_\text{el}$), the model places smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe ($\sim$8.5~GW$_\text{el}$) (see bottom left). Hydrogen prices vary strongly across the continent, with low price ``valleys'' where the CU locates with around 60~€/MWh, and high price ``peaks'' in Central Europe with around 95~€/MWh. Despite the transport system, carbon prices still vary from 110~€/t in Central Europe to 160~€/t at sequestration sites.

In the \hydrogenscenario{}, the main hydrogen transportation routes go from Spain (50~GW) and the United Kingdom (18~GW) to Central Europe to supply local carbonaceous fuel production with captured carbon from point sources (see Fig~\ref{fig:balance_map}, bottom right). Also, the \Htwo{} network supplies spatially-fixed \Htwo{} demand for industry in Central Europe (2.5~GW in total) at much lower costs than local electrolysis or SMR. Carbon sequestration primarily locates in Portugal where large DAC facilities operate (14~kt/h, see Fig~\ref{fig:balance_map}, top right). \carbon{} prices are low in Central Europe (110-120~€/t) and highest at sequestration regions (160~€/t). Hydrogen prices have flattened out at a relatively low level of €60~/MWh in all regions.

In both cases, CU is placed where material inputs, \carbon{} and \hydrogen{}, are provided at minimal cost, i.e., in price valleys of hydrogen and carbon in the \carbongrid{} and \hydrogenscenario{} respectively. In the \carbonscenario{}, CU co-locates with low-cost hydrogen production and uses transported carbon from the inland. In the \hydrogenscenario{}, CU co-locates with carbon point sources and uses transported low-cost hydrogen.
Both scenarios effectively combine low-cost renewable hydrogen production and carbon management for combustion and spatially-fixed industrial processes.

However, both transport systems provide individual cost-benefits to the system. While the carbon grid makes it possible to transport captured carbon for sequestration at low cost, the dependence on DAC near the coast causes higher costs in the hydrogen scenario{}. On the other hand, the \hydrogengrid{} is able to supply the spatially-fixed hydrogen demands in Central Europe with low-cost hydrogen imports, while the \modCO{} is forced to deploy electrolysis in regions with poor renewable resources as well as additional gas infrastructure in Central Europe. The latter aspect dominates the first, leading to lower system cost in the \hydrogenscenario{}.

% To understand the dynamics of the \modCO{}, we observe that the system locates CU in the regions where it can minimize costs for its material inputs, carbon and hydrogen. As the \COtwo{} network largely smoothes out carbon prices across regions, CU is located in the hydrogen price valleys on the periphery of the continent. Similarly, the model locates CS in regions where it can minimize the costs of the carbon to be sequestered. Thus, CS is located in the carbon price valleys among the subset of regions with access to offshore sequestration sites. Due to the \COtwo{} network these low-price regions are not strongly pronounced, but still prevalent and impact CS locations. DAC is only used for CU and thus co-located in regions with a mix of low electricity costs (for running DAC) and low hydrogen costs.

% To explain the dynamics of the \modH{}, we observe that the model locates CU in the carbon price valleys across Western, Central and Eastern Europe (see top right), as the \Htwo{} network smoothes out hydrogen prices across the entire continent (see bottom right). CS is again located in the carbon price valleys among those regions with access to offshore sequestration sites. Due to the lack of a \COtwo{} network the full CC potentials from point sources are exhausted in all offshore sequestration regions, and the carbon is then sequestered. The additionally needed carbon to reach 200~Mt of CS is captured via DAC in the regions with access to offshore sequestration sites with the lowest costs for DAC, with the largest share deployed in Portugal (see top right). DAC is also used for CU in this model, and for both purposes DAC is located in the regions with the lowest costs for electricity.

% The increase in captured carbon from point sources in the models with \carbon{} and/or \hydrogen{} transport is more related to an upgrading of existing industrial or power plant facilities to capture carbon rather than a shift in technologies. This is evident when analyzing the capture share of the different CC technologies (see Fig.~\ref{fig:captureshare_line}).

% In summary, the implementation of \COtwo{} and \Htwo{} networks is instrumental in unlocking CC potentials and reducing system costs. However, despite the \modCO{} demonstrating higher amounts and shares of CC from point sources and therefore less DAC deployment, the \modH{} achieves even  lower total system costs (see Fig.~\ref{fig:cost_bar}).

\begin{figure*}[ht!]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_carbon.png}
        \label{fig:balance_map_carbon_full}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_hydrogen.png}
        \label{fig:balance_map_hydrogen_full}
    \end{subfigure}
    \caption{Optimal operation, flows and prices of the carbon (left) and hydrogen (right) sectors for the \hybridmodel{} in the net zero scenario. For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Carbon Sequestration offshore are drawn as full circles. Lines and arrows show the interregional transportation. \carbon{} from point-source in the inland either supplies local CU with imported \hydrogen{} or facilitates sequestration in nearby offshore regions.
    % Carbon network looks the same as in~\cite{morbeeOptimisedDeploymentEuropean2012}: two backbones, one in the nothern Europe other in south east.
    }
    \label{fig:balance_map_full}
\end{figure*}

The \hybridscenario{}, the model combines the advantages of hydrogen and carbon networks, resulting in the highest system cost reduction of \label{}5.2\% (\label{}41~bn€/a) compared to the \modBase{}. As in the \hydrogenscenario{}, the model transports low-cost hydrogen to the point sources in Central Europe to supply CU, but also transports low-cost carbon from point sources close the shore to sequestration sites (see Fig.~\ref{fig:balance_map_full}). This reduces reliance on high-cost DAC and provides low-cost hydrogen across the regions.
